---
title: "Supplementary materials"
bibliography: 
  - references.bib
  - references2.bib
  - references_packages.bib
csl: ama.csl
pandoc_args: [       "-F", "./pandoc-word-pagebreak"     ]
output:
  bookdown::word_document2:
      toc: true
      toc_depth: 2
      reference_docx: word-styles-reference-01.docx
header-includes:
    - \usepackage[labelformat=empty]{caption}
---

``` {r, echo = FALSE, warning=FALSE, message=FALSE }
library(dplyr)
library(kableExtra)
library(rio)
library(MASS)
library(here)
library(epitools)
library(flextable)
library(ftExtra)

source(here("R","0_library.R"))
source(here("R","3_results.R"))

knitr::opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE)

```

\newpage

# Supplementary Tables

## Supplementary Table 1

```{r ft.align="left", tab.cap = "Change in data availability openness, stratified by preprint decision category and subsequently by journal open data policy."}
ft_s1_supp2
```

\newpage

## Supplementary Table 2

```{r ft.align="left", tab.cap = "Change in data availability statement openness, stratified by journal."}
ft_s1_supp
```

\newpage

## Supplementary Table 3

```{r tabs4, ft.aign = "left", tab.cap = "Categories for the 22 records that were not open despite being published in a journal that requires open data."}

ft_s4

```

\newpage

## Supplementary Table 4

```{r tabs2, ft.align="center", tab.cap='Comparison of code availability declarations between data availability statements and full text manuscripts.'}

ft_s2

```

\newpage

# Supplementary Materials

##  Supplementary Material 1

For items that met the criteria for multiple categories (e.g. "Most of the data analyzed in this manuscript are provided either within the manuscript itself, or in the manuscript posted by Sasani et al. on bioRxiv at  https://www.biorxiv.org/content/10.1101/552117v2 and its accompanying links; additional data may be accessed by contacting the corresponding author (Dr. Cawthon)."), we took a conservative approach, and categorized them based on the least permissive aspect of the statement. This approach was taken on the basis that __*all*__ data must be available - for example, in the above statement, the additional information that is available on request may be required in order to reproduce the analysis. 
